Soss

### 2020-01-05 paper/paper.tex
edits

\newmintinline[jl]{julia}{}
\begin{document}

@@

    \begin{minted}{julia}
m = @model I,J begin
    λ ~ HalfCauchy()
    σI ~ HalfNormal(λ)
    σJ ~ HalfNormal(λ)

@@

\begin{minted}{julia}
julia> d = m(I=2, J=3); rand(d) # or just rand(m(I=2, J=3))
(I = 2, J = 3, yhat = [-0.52 -1.5 0.0037; -0.81 -1.85 -0.28], σ = 2.5, λ = 1.3, σJ = 0.72, σI = 0.41, 
 β = [-0.54, -1.5, -0.023], α = [0.026, -0.26], y = [-0.45 -2.0 2.2; -3.4 -3.9 2.3])
\end{minted}

@@

\begin{minted}{julia}
struct Model{A,B,M} 
    args  :: Vector{Symbol}
    vals  :: NamedTuple
    dists :: NamedTuple
    retn  :: Union{Nothing, Symbol, Expr}
end
\end{minted}

The ability to generate arbitrary  Julia code makes it difficult to measure performance in Soss. When a performance bottleneck is found, developers or users can ask Soss to generate something different. This design places no a priori constraints on performance.

@@

% \begin{minted}{julia}
%   abstract type LCTerm end
%   struct Lam{Arg, Body <: LCTerm} <: LCTerm end
%   struct App{Fn<:LCTerm, Arg<:LCTerm} <: LCTerm end
%   struct Var{N} <: LCTerm end

@@

\begin{figure}[t]
\centering
\begin{subfigure}[t]{0.48\textwidth}
\begin{minted}{julia}
julia> Soss.sourceRand()(m)
quote
    σ = rand(HalfNormal())
    λ = rand(HalfCauchy())
    σJ = rand(HalfNormal(λ))
    σI = rand(HalfNormal(λ))
    β = rand(iid(J, Normal(0, σJ)))
    α = rand(iid(I, Normal(0, σI)))
    yhat = α .+ β'
    y = rand(For(I, J) do i, j
                Normal(yhat[i, j], σ)
            end)
    (I=I, J=J, yhat=yhat, λ=λ, σI=σI, 
    σJ=σJ, α=α, β=β, σ=σ, y=y) 
end
\end{minted}
\caption{\footnotesize{\texttt{rand}}}
\end{subfigure}
\begin{subfigure}[t]{0.48\textwidth}
\begin{minted}{julia}
julia> Soss.sourceLogpdf()(m)
quote
    _ℓ = 0.0
    _ℓ += logpdf(HalfNormal(), σ)
    _ℓ += logpdf(HalfCauchy(), λ)
    _ℓ += logpdf(HalfNormal(λ), σJ)
    _ℓ += logpdf(HalfNormal(λ), σI)
    _ℓ += logpdf(Normal(0, σJ) |> iid(J), β)
    _ℓ += logpdf(Normal(0, σI) |> iid(I), α)
    yhat = α .+ β'
    _ℓ += logpdf(For(I, J) do i, j
                Normal(yhat[i, j], σ)
            end, y)
    return _ℓ
end
\end{minted}
\caption{\footnotesize{\texttt{logpdf}}}
\end{subfigure}
\caption{Generated code when calling \footnotesize{\texttt{rand(m)}} and \footnotesize{\texttt{logpdf(m,x)}}.} 
\label{} 
\end{figure}

@@

\begin{minted}{julia}
julia> using Revise, Soss, Random

>>>>

\newmintinline[jl]{julia1}{}
\begin{document}

@@

    \begin{minted}{julia1}
m = @model I,J begin
    λ ~ HalfCauchy()
    σI ~ HalfNormal(λ)
    σJ ~ HalfNormal(λ)

@@

\begin{minted}{julia1}
julia> d = m(I=2, J=3); rand(d) # or just rand(m(I=2, J=3))
(I = 2, J = 3, yhat = [-0.52 -1.5 0.0037; -0.81 -1.85 -0.28], σ = 2.5, λ = 1.3, σJ = 0.72, σI = 0.41, 
 β = [-0.54, -1.5, -0.023], α = [0.026, -0.26], y = [-0.45 -2.0 2.2; -3.4 -3.9 2.3])
\end{minted}

@@

Soss's inference primitives generate code on demand using the \jl#GeneralizedGenerated.jl# library, which introduces a \jl#@gg# macro that works as an extension of Julia's built-in \jl#@generated# macro for \emph{generated functions} \cite{bezanson2012julia}. Similarly to \jl#@generated#, code generated using a \jl#@gg# function cannot depend on the values of its arguments, but only on their types.

This is problematic, since we clearly need code generation to depend on the model itself. To address this, \jl#GeneralizedGenerated# includes facilities for lifting Julia values to the type level.

Using this, the type of a Soss model is \jl#Model{A,B,M}#, where 
\begin{itemize}
    \item \jl#A# gives the type of the \emph{arguments} (\jl#A=NamedTuple{T}# for some tuple of symbols \jl#T#),
    \item \jl#B# is a type-level representation of the \emph{body}, and
    \item \jl#M# is a type-level representation of the \emph{module} in which the model was defined.
\end{itemize}

In this, the ``body'' is the \jl#begin...end# following the (possibly empty) arguments of a model. Knowing the module is important in case values from that module are referenced in the model.

\begin{figure}[t]
    \begin{subfigure}[b]{0.48\textwidth}
        \begin{minted}[autogobble]{julia1}
        struct Model{A,B,M} 
            args  :: Vector{Symbol}
            vals  :: NamedTuple
            dists :: NamedTuple
            retn  :: Union{Nothing, Symbol, Expr}
        end

        julia> m.dists
        ( λ = :(HalfCauchy())
        , σI = :(HalfNormal(λ))
        , σJ = :(HalfNormal(λ))
        , α = :(Normal(0, σI) |> iid(I))
        , β = :(Normal(0, σJ) |> iid(J))
        , σ = :(HalfNormal())
        , y = :(For(I, J) do i, j
                Normal(yhat[i, j], σ)
                end))
        \end{minted}

    \end{subfigure}    
    \begin{subfigure}[b]{0.48\textwidth}
        \begin{minted}[autogobble]{julia1}
            julia> m.vals
            (yhat = :(α .+ β'),)

            julia> typeof(m)
            Model{NamedTuple{(:I, :J),T} where T<:Tuple,
                TypeEncoding(begin
                    σ ~ HalfNormal()
                    λ ~ HalfCauchy()
                    σJ ~ HalfNormal(λ)
                    σI ~ HalfNormal(λ)
                    β ~ Normal(0, σJ) |> iid(J)
                    α ~ Normal(0, σI) |> iid(I)
                    yhat = α .+ β'
                    y ~ For(I, J) do i, j
                            Normal(yhat[i, j], σ)
                        end
                end),TypeEncoding(Main)}
        \end{minted}
    \end{subfigure}    
    \caption{The definition of the \texttt{Model} type, and some details of the \texttt{m} model from Figure \ref{fig:model}.}
\end{figure}

The flexibility of code generation in Soss makes it difficult to measure performance. When a performance bottleneck is found, developers or users can ask Soss to generate something different. This design places no a priori constraints on performance.

@@

% \begin{minted}{julia1}
%   abstract type LCTerm end
%   struct Lam{Arg, Body <: LCTerm} <: LCTerm end
%   struct App{Fn<:LCTerm, Arg<:LCTerm} <: LCTerm end
%   struct Var{N} <: LCTerm end

@@

\begin{figure}[t]
\centering
\begin{subfigure}[t]{0.48\textwidth}
\begin{minted}{julia1}
julia> Soss.sourceRand()(m)
quote
    σ = rand(HalfNormal())
    λ = rand(HalfCauchy())
    σJ = rand(HalfNormal(λ))
    σI = rand(HalfNormal(λ))
    β = rand(iid(J, Normal(0, σJ)))
    α = rand(iid(I, Normal(0, σI)))
    yhat = α .+ β'
    y = rand(For(I, J) do i, j
                Normal(yhat[i, j], σ)
            end)
    (I=I, J=J, yhat=yhat, λ=λ, σI=σI, 
    σJ=σJ, α=α, β=β, σ=σ, y=y) 
end
\end{minted}
\caption{\footnotesize{\texttt{rand}}}
\end{subfigure}
\begin{subfigure}[t]{0.48\textwidth}
\begin{minted}{julia1}
julia> Soss.sourceLogpdf()(m)
quote
    _ℓ = 0.0
    _ℓ += logpdf(HalfNormal(), σ)
    _ℓ += logpdf(HalfCauchy(), λ)
    _ℓ += logpdf(HalfNormal(λ), σJ)
    _ℓ += logpdf(HalfNormal(λ), σI)
    _ℓ += logpdf(Normal(0, σJ) |> iid(J), β)
    _ℓ += logpdf(Normal(0, σI) |> iid(I), α)
    yhat = α .+ β'
    _ℓ += logpdf(For(I, J) do i, j
                Normal(yhat[i, j], σ)
            end, y)
    return _ℓ
end
\end{minted}
\caption{\footnotesize{\texttt{logpdf}}}
\end{subfigure}
\caption{Generated code when calling \footnotesize{\texttt{rand(m)}} and \footnotesize{\texttt{logpdf(m,x)}}.} 
\label{} 
\end{figure}

@@

\begin{minted}{julia1}
julia> using Revise, Soss, Random
